visdat: Visualising Whole Data Frames

### 2017-08-16 paper/paper.Rmd
update JOSS paper and adjust figures

---
title: 'visdat: Visualising whole data frames'
tags:
  - visualisation
  - R
  - exporatory data analysis
authors:
 - name: Nicholas Tierney
   orcid: 0000-0003-1460-8722
   affiliation: 1
affiliations:
 - name: Queensland University of Technology
   index: 1
date: 09 January 2017
bibliography: paper.bib
output: 
  html_document: 
    keep_md: yes
---

@@

When you get a new data set, you need to look at the data to get a sense of what it contains and potential problems with it. You read in your data, but you then need to look at it to understand what you need to do to make it ready for analysis. A phrase that often arises in this process is "looking at the data". It is helpful to look more closely at what this means. You can look at the first six rows of data, the head of the data:

```{r head-iris}

head(iris)

@@

Another alterantive is `glimpse`, from the `dplyr` package [@dplyr]

```{r glimpse}

dplyr::glimpse(iris)

This has a better printing method that gives us more information about the data, displaying doubles and a factor. However, we don't always have data like the canonical iris dataset. Let's take a look at some data that might be a bit more typical of "messy" data.

```{r visdat-glimpse}

library(visdat)
dplyr::glimpse(typical_data)

Looking at this, you might then ask:

> Isn't it odd that Income is a factor? And Age is a character?

And you might start to wonder what else is different, what else changed?

And it might be a bit unclear where to go from there. Do you plot the data? Why does my plot look wierd? What are these other strange features in the data? The `visdat` package provides visualisations of an entire dataframe at once. Initially inspired by [`csv-fingerprint`](https://github.com/setosa/csv-fingerprint), `visdat` provides tools to create heatmap-like visualisations of an entire dataframe. `visdat` is an R [@Rcore] package provides 2 main functions `vis_dat` and `vis_miss`.

`vis_dat()` helps explore the data class structure and missingness:

```{r load-data}

vis_dat(typical_data)

And the `vis_miss` function provides a custom plot for missing data.

```{r}

vis_miss(typical_data)

These functions provide useful tools to help "get a look at the data", using principled visualisation techniques. The plots are built using ggplot2 [@ggplot2], which provides a consistent and powerful framework for visualisations. This means that users can customise and extend graphics from visdat very easily.

>>>>

---
title: 'visdat: Visualising Whole Data Frames'
authors:
- affiliation: 1
  name: Nicholas Tierney
  orcid: 0000-0003-1460-8722
date: "01 August 2017"
output:
  pdf_document: default
  html_document:
    keep_md: yes
bibliography: paper.bib
tags:
- visualisation
- R
- exporatory data analysis
affiliations:
- index: 1
  name: Monash University
---

JOSS requirements:

- A summary describing the high-level functionality and purpose of the software for a diverse, non-specialist audience
- A clear statement of need that illustrates the purpose of the software
- A list of key references including a link to the software archive
- Mentions (if applicable) of any ongoing research projects using the software or recent scholarly publications enabled by it

@@

When you receive a new dataset you need to look at the data to get a sense of what is in it and understand potential problems and challenges to get it analysis-ready. "Taking a look at the data" can mean different things. For example, examining statistical summaries (minimum, maximum, mean, interquartile range), finding missing values, checking data formatting, creating graphical summaries such as histograms, scatterplots, boxplots, and more.

When handling typical "messy" data, these preliminary exploratory steps can become difficult to perform when you encounter data that is not what you expect. For example, income might be a factor instead of numeric, date could be a number not a character string or a date class, or values could be missing when they shouldn't be. Often times you realise that you have expectations of the data, but it can be hard to know what they are exactly until they are a problem. This is sort of like how you might not think to buy more lightbulbs until one goes out - when you go to use the data in an exploratory scatter plot, or in a preliminary model, you don't always realise your data is in the wrong format until that moment. What you need is a birds eye view of the data that tells you what classes are in the dataframe, and where the missing data are.

`visdat` is an R [@Rcore] package that provides a tool to "get a look at the data", creating heatmap-like visualisations of an entire dataframe - classes in the data, missing values, and comparisons between two datasets. `visdat` takes inspiration from [`csv-fingerprint`](https://github.com/setosa/csv-fingerprint), and is powered by ggplot2 [@ggplot2], which provides a consistent and powerful framework for visualisations that can be extended if needed.

Plots are presented in an intuitive way, reading top down, just like your data. Below is a plot of some example typical data containing missing values and data of a variety of classes.

```{r load-data}
library(visdat)
vis_dat(typical_data)

@@

`visdat` will continue to be improved over time, to improve speed in computation and also add interactive plots.

# Acknowlegements

I would like to thank the two reviewers, Mara Averick and Sean Hughes, for their helpful suggestions that resulted in a much better package, and rOpenSci for providing the support of the onboarding package review that facilitated these improvements.
